---
title: "Exercise 03"
author: "Nianzhen Gu (NianzhenGu)"
date: "08 10 2023"
format:
  html:
    self-contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory Data Analysis

Do an exploratory data analysis of a matrix of expression values. The data consists of expression values for samples that were treated with DMSO and TSA. The samples were measured using three technologies: bulk, IFC96, IFC800. See the two RDS files `counts.RDS` and `phenodata.RDS`.

The experiment has two experimental factors "Treatment" and "Technology". Treatment has two levels, Technology has three levels. Samples that have the same combination of Treatment & Technology are replicates.

## Data Import

```{r }
x = readRDS("counts.RDS")
anno = readRDS("phenodata.RDS")
head(anno)

```

## Compute and visualize basic statistics on the experimental design

How many replicates are there for each combination of factor levels? How many genes have an expression value above 0 in each sample? Are there systematic differences between the samples in the different conditions (treatment & technology). Visualize the following statistics per sample:

How many replicates are there for each combination of factor levels?
```{r}
comb_max <- matrix(0, nrow = 3, ncol = 2)
rownames(comb_max) <- c("bulk","IFC96","IFC800")
colnames(comb_max) <- c("DMSO","TSA")

for (i in 1:nrow(anno)) {
  treatment <- anno[i,1]
  technology <- anno[i,2]
  
  comb_max[technology, treatment] <- comb_max[technology, treatment]+1
}

print(comb_max)
```

How many genes have an expression value above 0 in each sample?
```{r}
sample_name <- colnames(x)
num_genes_above_zero <- apply(x > 0, 2, sum)
cat("Number of genes with expression values above 0 in each sample:\n")
for (i in seq_along(num_genes_above_zero)) {
  cat(sample_name[i], ":", num_genes_above_zero[i], "\n")
}
```


Are there systematic differences between the samples in the different conditions (treatment & technology)？

From the above values, we can see that there are differences between the samples in the different conditions in terms of gene expression values. 

number of detected genes (Color the samples by the experimental condition.)
```{r}
library(ggplot2)

sample_condition_table <- data.frame(Sample_Names = sample_name,
                      Number_of_Genes = 0,
                      Sum_of_Reads = 0) # create a dataframe to store the number of zeros in each sample

for (sample in sample_name) {
  category <- paste(anno[sample, "Treatment"],
                    anno[sample, "Technology"],
                    sep = "&")
  count_zero <- sum(x[, sample] == 0) # count 0 in each sample
  num_genes <- nrow(x) - count_zero # genes number, for later use
  sum_reads <- sum(x[, sample]) # sum of reads, for later use
  
  sample_condition_table[sample_condition_table$Sample_Names==sample, "Number_of_Genes"] <- num_genes
  sample_condition_table[sample_condition_table$Sample_Names==sample, "Sum_of_Reads"] <- sum_reads
  sample_condition_table[sample_condition_table$Sample_Names==sample, "Category"] <- category
}

head(sample_condition_table)

ggplot(sample_condition_table, aes(x = Sample_Names, y = Number_of_Genes, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of detected genes", x = "Sample", y = "Number of genes") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=4))

```

total sum of counts per sample (serves as a proxy for the number of sequenced reads per sample) (Color the samples by the experimental condition.)
```{r}
ggplot(sample_condition_table, aes(x = Sample_Names, y = Sum_of_Reads, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "Total sum of counts per sample", x = "Sample", y = "Total sum of counts") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=4))

```


## Normalize the data

Scale the columns so that the total sum of all columns are identical

```{r}
# install.packages("BiocManager")
# library(BiocManager)
# BiocManager::install("preprocessCore")

library(preprocessCore)
normalized_data <- normalize.quantiles(x, keep.names = TRUE)
```

## Transform the data to log-scale

Use the function `log1p` to transform the data to log-scale

```{r}
# Apply log1p transformation to each element
log_scaled_data <- log1p(normalized_data)

```

## Visualize the distribution of the expression values

Use violin plots and boxplots to visualize the distribution of the expression values. 
Aggregate the replicates in the same experimental group into one average sample, and plot the averaged sample. Color by experimental factors.

```{r}
categories <- unique(sample_condition_table$Category)
gene_names <- rownames(x)

# Filter the table based on the specified categories
filtered_data <- sample_condition_table[sample_condition_table$Category %in% categories, ]

# Get the names for each category
result <- split(filtered_data$Sample_Names, filtered_data$Category)


average_matrix <- matrix(0, nrow = length(gene_names), ncol = length(categories), dimnames = list(gene_names, categories))


for (category in categories) {
  sum <- rowSums(log_scaled_data[, result[[category]]])
  average_matrix[, category] <- sum / length(result[[category]])
}

head(average_matrix)

# Load the required packages
library(ggplot2)
library(tidyr)
library(patchwork)

average_df <- as.data.frame(average_matrix)
average_df_long <- gather(average_df, key = "category", value = "value")

# Create a violin plot
violin_plot <- ggplot(average_df_long, aes(x = category, y = value)) +
  geom_violin(fill = "skyblue", color = "black") + 
  labs(title = "Violin Plot") +
  theme_minimal()

# Create a box plot
box_plot <- ggplot(average_df_long, aes(x = category, y = value)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(title = "Box Plot") +
  theme_minimal()

violin_plot

box_plot

```

## Most variable genes

Identify the 500 most variable genes (with largest variance across samples) and continue working with those
```{r}
variance <- apply(log_scaled_data, 1, var)
sorted_variances <- sort(variance, decreasing = TRUE)
top500 <- sorted_variances[1:500]
print(names(top500))
```

## Sample correlations

Compute and visualize the sample-to-sample correlations

```{r}
average_df2 <- as.data.frame(average_df[names(top500), ])
correlations <- cor(average_df2, method = "pearson")

# Visualization
library(reshape2)
correlation_plot <- ggplot(data = melt(correlations),
             aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "tomato") +
  labs(title = "Sample-to-sample correlations",
       x = "Sample",
       y = "Sample",
       fill = "Correlation Value") +
  theme(axis.text.x = element_text(angle = 300))
correlation_plot
```

## Clustering

Compute and visualize a hierarchical clustering of the samples, use the method `hclust`
```{r}
clustering <- hclust(dist(t(average_df2)))
plot(clustering, main = "Hierarchical clustering",
     xlab = "Samples", ylab = "Distance",
     labels = colnames(average_df2),
     hang = -1, col = "steelblue", cex = 0.8)
```

## Heatmap

Use the package `pheatmap` to generate a heatmap of the expression data.

```{r}
library(pheatmap)
pheatmap(average_df2, main = "Heatmap of the expression data", show_rownames = FALSE)

```

